---
title: "Latent Class Analysis of Meritocratic Beliefs"
subtitle: "Data analysis"
author:
  - name: Andreas Laffert
    corresponding: true
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 1
  - name: Juan Carlos Castillo
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
  - name: Tomás Urzúa
    orcid: 0009-0008-3167-6819
    email: tomas.urzua@ug.chile.cl
    affiliations:
      - ref: 1    
  - name: René Canales
    orcid: 0009-0002-4299-2725
    email: renecanales@uc.cl
    affiliations:
      - ref: 1  
affiliations: 
  - id: 1
    name: Department of Sociology, University of Chile
date: today
lang: en
fontsize: 12pt
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-location: right
    toc-depth: 2
    toc-expand: 2
    toc-title: Contents
    number-sections: true
    number-depth: 3
    theme:
      - cosmo
      - edumer_html.scss
    code-link: true
    title-block-banner: true
editor_options: 
  chunk_output_type: console
bibliography: "../input/bib/jusmer.bib"
csl: "../input/bib/apa6.csl"
---


```{r}
#| label: set
#| echo: false
#| message: false
#| warning: false

library(knitr)
knitr::opts_chunk$set(echo = TRUE, include = TRUE, warning = FALSE, message = FALSE)

table_format <- if(is_html_output()) {
  "html"
} else if(is_latex_output()) {
  "latex"
}
table_format2 <- if(is_html_output()) {
  T
} else if(is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```

# Aim

This study applies latent class analysis (LCA) to a multidimensional scale of meritocratic beliefs developed by  @castillo_multidimensional_2023 to identify distinct latent profiles of how individuals combine meritocratic and non-meritocratic beliefs.

Rather than focusing on validating the measurement model, the analysis seeks to uncover patterns of response configurations across items, thereby identifying underlying typologies of meritocratic orientations.

The dataset used is `db_proc.RData` generated in the `prod_prep.R` script.

# Data

- Dataset: EDUMERCO (2025)
- Mode: Online survey (CAWI)
- Population: Adults in the Metropolitan Region of Chile
- Sampling: Non-probability quota-based design
  - Quotas based on age, gender, education, and socioeconomic status
- Sample size: N = 3,470
- Analytical sample: N = 2,740

# Measures

## Meritocracy scale

We use a multidimensional scale that distinguishes between (@tbl-merit):

1. Perceptions and preferences
2. Meritocratic and non-meritocratic principles

This results in four dimensions:

- Perceived meritocracy: beliefs about whether effort and ability are rewarded
- Perceived non-meritocracy: beliefs about the role of connections and family wealth
- Preference for meritocracy: normative support for rewarding effort and talent
- Preference for non-meritocracy: normative acceptance of advantages linked to privilege

Each dimension is measured with two items, for a total of eight items. Responses are recorded on a four-point Likert scale ranging from *strongly disagree* (1) to *strongly agree* (4).



```{r}
#| tbl-cap: Items for perceptions and preferences for meritocracy scale
#| tbl-cap-location: top
#| echo: false
#| label: tbl-merit

 # remotes::install_github('haozhu233/kableExtra@a6af5c0')  # for problems with collapse_rows()
 pacman::p_load(knitr, kableExtra, dplyr, here)
 table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }
 tabitems <- read.csv(file = here("input/table-01.csv"),header = 1,sep = ",",encoding = "UTF-8") # call file  generated externally for the table
 cnames <- c("Components", "Dimensions","Item (English)" , "Item original (Spanish)")

 knitr::kable(tabitems, "html", booktabs = T, linesep = "",col.names = cnames) %>%
   kableExtra::kable_styling(
     full_width = T,
     latex_options = c("hold_position"),
     position = "center",
     font_size = 9.5,
     bootstrap_options=c("striped", "bordered")) %>%
   kableExtra::column_spec(column = 1, width = "2cm", ) %>%
   kableExtra::column_spec(column = 2,width = "2 cm") %>%
   kableExtra::collapse_rows(columns = 1:2,valign = "middle")
```


# Method

## Latent class analysis

Latent class analysis (LCA) is a statistical method that models associations among observed variables by assuming that an unobserved categorical latent variable explains their underlying structure. This latent variable represents discrete subgroups within the population, where individuals in the same class share similar response patterns.

LCA is especially appropriate for categorical and ordinal indicators, such as Likert-scale items.

### Analytical purpose

LCA can serve both exploratory and confirmatory purposes.

- As an exploratory technique, it identifies latent typologies of belief configurations.
- As a confirmatory technique, it allows researchers to test hypotheses about heterogeneity in the population.

Substantively, LCA enables the derivation of empirically grounded ideal types of meritocratic orientations.

### Key assumptions

- Each individual belongs to one latent class.
- Individuals within the same class share similar response probabilities.
- Observed indicators are conditionally independent once class membership is taken into account (local independence).

### Model parameters

The model estimates two main sets of parameters:

1. Class probabilities  
   These indicate the probability of belonging to each latent class and reflect the relative size of each group in the population.

2. Conditional response probabilities  
   These indicate the probability of endorsing a particular response category, given membership in a latent class. These parameters are used to interpret and label the classes.

## Interpretation

Latent classes are defined by distinct response profiles across the eight items. These profiles capture different combinations of:

- meritocratic and non-meritocratic beliefs
- perceptions and normative preferences

This allows us to identify complex belief systems in which individuals may, for example, endorse meritocratic ideals while simultaneously recognizing, or even accepting, non-meritocratic advantages.

## Model estimation and fit

Models are estimated iteratively by increasing the number of classes.

Model selection is based primarily on standard fit criteria, especially:

- Bayesian Information Criterion (BIC)
- Akaike Information Criterion (AIC)

Lower values indicate a better fit, while also balancing model complexity and parsimony.

## Contribution

This strategy shifts the analysis from a variable-centered approach to a person-centered one. In doing so, it allows us to:

- capture heterogeneity in meritocratic belief systems
- identify meaningful subgroups of respondents
- examine how meritocratic and non-meritocratic beliefs coexist within individuals



# Libraries

```{r}
#| label: packages
#| include: true

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse,
               sjmisc, 
               here,
               lavaan,
               psych,
               corrplot,
               ggdist,
               sjlabelled,
               patchwork,
               RColorBrewer,
               poLCA,
               reshape,
               summarytools)

options(scipen=999)
rm(list = ls())
```

# Data

```{r}
#| label: data
#| echo: true

load(here("input/data/proc/db_proc.RData"))

glimpse(db)

db <- db %>% 
  mutate(
    income_4 = case_when(
      income %in% c(
        "Menos de $280.000 mensuales liquidos",
        "De $280.001 a $380.000 mensuales liquidos",
        "De $380.001 a $470.000 mensuales liquidos"
      ) ~ "Low",
      
      income %in% c(
        "De $470.001 a $610.000 mensuales liquidos",
        "De $610.001 a $730.000 mensuales liquidos"
      ) ~ "Middle-low",
      
      income %in% c(
        "De $730.001 a $890.000 mensuales liquidos",
        "De $890.001 a $1.100.000 mensuales liquidos"
      ) ~ "Middle-high",
      
      income %in% c(
        "De $1.100.001 a $2.700.000 mensuales liquidos",
        "De $2.700.001 a $4.100.000 mensuales liquidos",
        "Mas de $4.100.001 mensuales liquidos"
      ) ~ "High",
      
      TRUE ~ NA_character_
    ),
    income_4 = factor(income_4, levels = c("Low", "Middle-low", "Middle-high", "High"))
  )


db$income_4 <- sjlabelled::set_label(db$income_4, "Income")

db$age <- sjlabelled::set_label(db$age, "Age")

db$sex <- if_else(db$sex == 1, "Male", "Female")

db$sex <- sjlabelled::set_label(db$sex, "Sex")

db$educ <- sjlabelled::set_label(db$educ, "Educational level")


db$pol <- sjlabelled::set_label(db$pol, "Political identification")



```

# Analysis

## Descriptives

```{r}
#| label: tbl-summary1
#| tbl-cap: "Descriptive statistics"
#| tbl-cap-location: top

vars_m <- c("perc_effort",
            "perc_talent",
            "perc_rich_parents",
            "perc_contact",
            "pref_effort",
            "pref_talent",
            "pref_rich_parents",
            "pref_contact")

t1 <- db %>% 
   dplyr::select(all_of(vars_m), age, sex, educ, income_4, pol)


df<-dfSummary(t1,
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               graph.magnif = 0.75,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = T,    # missing
               graph.col = T, # plot
               valid.col = T, # n valido
               col.widths = c(20,10,10,10,10,10))

df$Variable <- NULL # delete variable column

print(df, method="render")
```


```{r}
#| label: fig-likert
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Distribution of responses on the scale of perceptions and preferences regarding meritocracy"
#| fig-cap-location: top

db <- db %>% 
  mutate(
    across(
      .cols = all_of(vars_m),
      .fns = ~as.numeric(.)
    ))
 
labels1 <- c("Strongly desagree" = 1, 
             "Desagree" = 2, 
             "Agree" = 3, 
             "Strongly agree" = 4)
db <- db %>% 
  mutate(
    across(
      .cols = all_of(vars_m),
      .fns = ~  sjlabelled::set_labels(., labels = labels1)
    )
  )


df <- db %>% 
  dplyr::select(all_of(vars_m)) %>% 
  drop_na()

theme_set(theme_ggdist())
colors <- RColorBrewer::brewer.pal(n = 4, name = "RdBu")


a <- df %>% 
  dplyr::select(perc_effort, 
                perc_talent,
                perc_rich_parents,
                perc_contact) %>% 
  sjPlot::plot_likert(geom.colors = colors,
                      title = c("a. Perceptions"),
                      geom.size = 0.8,
                      axis.labels = c("Effort", "Talent", "Rich parents", "Contacts"),
                      catcount = 4,
                      values  =  "sum.outside",
                      reverse.colors = F,
                      reverse.scale = T,
                      show.n = FALSE,
                      show.prc.sign = T
  ) +
  ggplot2::theme(legend.position = "none")

b <- df %>% 
  dplyr::select(pref_effort, 
                pref_talent,
                pref_rich_parents,
                pref_contact) %>% 
  sjPlot::plot_likert(geom.colors = colors,
                      title = c("b. Preferences"),
                      geom.size = 0.8,
                      axis.labels = c("Effort", "Talent", "Rich parents", "Contacts"),
                      catcount = 4,
                      values  =  "sum.outside",
                      reverse.colors = F,
                      reverse.scale = T,
                      show.n = FALSE,
                      show.prc.sign = T
  ) +
  ggplot2::theme(legend.position = "bottom")

likerplot <- a / b + plot_annotation(caption = paste0("Source: own elaboration based on Survey EDUMERCO"," (n = ",dim(df)[1],")"
))

likerplot

```

```{r}
#| label: fig-cor
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Correlation matrix"
#| fig-cap-location: top

M <- df %>% 
  psych::polychoric()

diag(M$rho) <- NA

rownames(M$rho) <- c("A. Perception Effort",
                     "B. Perception Talent",
                     "C. Perception Rich parents",
                     "D. Perception Contacts",
                     "E. Preference Effort",
                     "F. Preference Talent",
                     "G. Preference Rich parents",
                     "H. Preference Contacts")

#set Column names of the matrix
colnames(M$rho) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)",
                    "(H)")

testp <- cor.mtest(M$rho, conf.level = 0.95)

#Plot the matrix using corrplot
corrplot::corrplot(M$rho,
                   method = "color",
                   addCoef.col = "black",
                   type = "upper",
                   tl.col = "black",
                   col = colorRampPalette(c("#E16462", "white", "#0D0887"))(12),
                   bg = "white",
                   na.label = "-") 

```


## Latent class analysis with dichotomized items

```{r}
#| include: false
# --- helpers -------------------------------------------------------------

get_K <- function(fit) {
  # robust: number of classes = columns of posterior
  if (!is.null(fit$posterior)) return(as.integer(ncol(fit$posterior)))
  # fallback: rows of probs for first item
  if (!is.null(fit$probs) && length(fit$probs) > 0) return(as.integer(nrow(fit$probs[[1]])))
  stop("Cannot determine K from poLCA object (no posterior/probs).")
}

get_npar <- function(fit) {
  if (!is.null(fit$npar)) return(as.numeric(fit$npar))
  # fallback if npar missing (rare): compute from probs + class shares
  K <- get_K(fit)
  Rj <- vapply(fit$probs, function(m) if (is.matrix(m)) ncol(m) else NA_integer_, integer(1))
  if (anyNA(Rj)) stop("Cannot compute npar: missing category counts in fit$probs.")
  as.numeric(sum(K * (Rj - 1)) + (K - 1))
}

get_n_patterns_obs <- function(fit) {
  # poLCA usually stores predicted cell counts for observed patterns as a vector
  if (!is.null(fit$predcell)) return(as.integer(length(fit$predcell)))
  # fallback: sometimes observed pattern table exists
  if (!is.null(fit$observed) && is.matrix(fit$observed)) return(as.integer(nrow(fit$observed)))
  NA_integer_
}

n_patterns_from_data <- function(data_indicators) {
  # data_indicators = db_lca with only indicators (your a1..a8), already NA-omitted
  as.integer(nrow(dplyr::distinct(data_indicators)))
}


entropy_stats <- function(fit) {
  P <- fit$posterior
  P <- pmax(P, 1e-12)
  N <- nrow(P); K <- ncol(P)
  
  # relative entropy (0..1): 1 = best separation
  rel_entropy <- 1 - (-sum(P * log(P)) / (N * log(K)))
  
  class_hat <- max.col(P)
  max_p <- apply(P, 1, max)
  
  avg_diag <- vapply(1:K, function(k) {
    idx <- which(class_hat == k)
    if (length(idx) == 0) NA_real_ else mean(P[idx, k])
  }, numeric(1))
  
  tibble(
    entropy = rel_entropy,
    mean_max_p = mean(max_p),
    p10_max_p = unname(quantile(max_p, 0.10)),
    avg_diag_min = suppressWarnings(min(avg_diag, na.rm = TRUE)),
    avg_diag_mean = mean(avg_diag, na.rm = TRUE)
  )
}

min_class_share <- function(fit) {
  Ppost <- fit$posterior
  N <- nrow(Ppost); K <- ncol(Ppost)
  class_hat <- max.col(Ppost)
  
  tibble(
    min_class_prior = min(fit$P),
    min_class_map   = min(tabulate(class_hat, nbins = K)) / N
  )
}

```


```{r}
#| include: false
# 3.3 LCA: dichotomized items ----

df <- db %>% 
  dplyr::select(id, ends_with("_d")) %>% 
  na.omit()

a1 <- df$perc_effort_d
a2 <- df$perc_talent_d
a3 <- df$perc_rich_parents_d
a4 <- df$perc_contact_d
a5 <- df$pref_effort_d
a6 <- df$pref_talent_d 
a7 <- df$pref_rich_parents_d
a8 <- df$pref_contact_d
id <- df$id

db_lca <- data.frame(id, a1, a2, a3, a4, a5, a6, a7, a8) %>% as_tibble()

db_lca <- db_lca %>% 
  mutate_at(.vars = 2:9, ~as.factor(.))

f <-cbind(a1, a2, a3, a4, a5, a6, a7, a8)~1

fit_lca <- function(K, data, formula, nrep = 50, maxiter = 3000, seed = 123) {
  set.seed(seed)
  poLCA(formula, data = data, nclass = K,
        nrep = nrep, maxiter = maxiter,
        verbose = FALSE, graphs = FALSE)
}

Ks <- 1:5
#fits <- lapply(Ks, fit_lca, data = db_lca, formula = f, nrep = 30)
#save(fits, file = here("output/fits_models_lca.RData"))

load(file = here("output/fits_models_lca.RData"))

fit_tbl <- data.frame(
  K    = Ks,
  logLik = sapply(fits, `[[`, "llik"),
  AIC  = sapply(fits, `[[`, "aic"),
  BIC  = sapply(fits, `[[`, "bic"),
  Gsq  = sapply(fits, `[[`, "Gsq"),
  Chisq = sapply(fits, `[[`, "Chisq")
)

fit_tbl
```

We estimated latent class models with one to five classes. Model fit improved monotonically as the number of classes increased, as indicated by decreasing AIC and BIC values and statistically significant likelihood-ratio tests at each step (@tbl-fit).

```{r}
#| label: tbl-fit
#| tbl-cap: "Fit comparision of latent class models"
#| tbl-cap-location: top

# ---- build table (no list-cols) -----------------------------------------

N_used <- nrow(db_lca)

m_patterns <- n_patterns_from_data(db_lca)  # scalar

# Robust extractors (avoid list-columns)
get_K <- function(fit) as.integer(ncol(fit$posterior))  # classes = cols of posterior
get_ll  <- function(fit) as.numeric(fit$llik)
get_aic <- function(fit) as.numeric(fit$aic)
get_bic <- function(fit) as.numeric(fit$bic)
get_npar <- function(fit) {
  if (!is.null(fit$npar)) return(as.numeric(fit$npar))
  # fallback: compute from probs
  K <- get_K(fit)
  Rj <- vapply(fit$probs, function(m) ncol(m), integer(1))
  as.numeric(sum(K * (Rj - 1)) + (K - 1))
}

# Table of fit indices
fit_tbl <- tibble(
  K     = map_int(fits, get_K),
  N     = nrow(db_lca),
  logLik= map_dbl(fits, get_ll),
  npar  = map_dbl(fits, get_npar),
  AIC   = map_dbl(fits, get_aic),
  BIC   = map_dbl(fits, get_bic)
) %>%
  arrange(K) %>%
  mutate(
    # Differences relative to preceding (K-1) model
    dAIC = AIC - lag(AIC),
    dBIC = BIC - lag(BIC),
    
    # Likelihood-ratio test (K vs K-1)
    # Test statistic: 2*(LL_K - LL_{K-1}) ~ Chi-square with df = npar_K - npar_{K-1}
    # Note: in LCA, models are not strictly nested in the regular sense;
    # this LRT is widely used as a heuristic, but interpret with caution.
    LRT = 2 * (logLik - lag(logLik)),
    df_LRT = npar - lag(npar),
    p_LRT = ifelse(!is.na(LRT) & df_LRT > 0, pchisq(LRT, df = df_LRT, lower.tail = FALSE), NA_real_)
  )


fit_tbl_report <- fit_tbl %>%
  transmute(
    K, N,
    logLik = round(logLik, 1),
    npar = npar,
    AIC = round(AIC, 1),
    dAIC = round(dAIC, 1),
    BIC = round(BIC, 1),
    dBIC = round(dBIC, 1),
    LRT = round(LRT, 1),
    df_LRT = df_LRT,
    p_LRT = signif(p_LRT, 3)
  )

fit_tbl_report %>% 
 kableExtra::kable(format = "html",
                    align = "c",
                    booktabs = T,
                    escape = F,
                    caption = NULL) %>%
  kableExtra::kable_styling(full_width = T,
                            latex_options = "hold_position",
                            bootstrap_options=c("striped", "bordered", "condensed"),
                            font_size = 23) %>%
  kableExtra::column_spec(c(1,8), width = "3.5cm") %>% 
  kableExtra::column_spec(2:7, width = "4cm") %>% 
  kableExtra::column_spec(4, width = "5cm")

```

```{r}
#| label: fig-visaic
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Visual: AIC/BIC by K (lower is better)"
#| fig-cap-location: top


# Visual: AIC/BIC by K (lower is better)
fit_long <- fit_tbl %>%
  dplyr::select(K, AIC, BIC) %>%
  pivot_longer(-K, names_to = "criterion", values_to = "value")

ggplot(fit_long, aes(x = K, y = value, group = criterion)) +
  geom_line() +
  geom_point() +
  facet_wrap(~criterion, scales = "free_y") +
  labs(x = "Number of classes (K)", y = "Value") 


```

```{r}
#| label: fig-dfvisaic
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "ΔAIC and ΔBIC (negative = improvement over K-1)"
#| fig-cap-location: top

# Visual: ΔAIC and ΔBIC (negative = improvement over K-1)
delta_long <- fit_tbl %>%
  dplyr::select(K, dAIC, dBIC) %>%
  pivot_longer(-K, names_to = "delta", values_to = "value")

ggplot(delta_long, aes(x = K, y = value, group = delta)) +
  geom_hline(yintercept = 0) +
  geom_line() +
  geom_point() +
  facet_wrap(~delta, scales = "free_y") +
  labs(x = "K", y = "Change vs K-1 (negative = better)") 
```



However, the incremental improvement in fit became substantially smaller after the four-class solution. Whereas the transition from one to four classes yielded large reductions in AIC and BIC, the five-class model provided only a modest additional improvement (AIC = 20,249.6; BIC = 20,509.9) relative to the four-class solution (AIC = 20,395.7; BIC = 20,602.7). Thus, although the five-class model fit the data slightly better in statistical terms, the gain was comparatively limited.

We therefore retained the four-class solution as the final model. This choice was guided by both statistical and substantive criteria: the four-class model achieved a strong balance between goodness of fit, parsimony, and interpretability, whereas the five-class solution did not yield enough additional differentiation to justify the increase in complexity.

## Four latent class model

The @tbl-prevalence shows the conditional probabilities of each item associated with a latent class and their respective sizes. The four-class solution suggests that respondents are not primarily differentiated along a simple continuum from “more meritocratic” to “less meritocratic.” Instead, the latent classes are organized by distinct combinations of: (1) perceptions of meritocracy, (2) perceptions of non-meritocratic advantage, and (3) normative preferences regarding whether merit or privilege should determine success. As such, the results reveal heterogeneity not only in empirical diagnoses of how society works, but also in normative evaluations of whether these mechanisms are considered acceptable. This interpretation is consistent with the item-response profiles across classes.

Two broader patterns structure the class solution. First, perceptions of non-meritocratic advantage are high across most classes, particularly regarding the role of wealthy parents and personal connections, suggesting that recognition of privilege is widespread. Second, the main line of differentiation lies in normative preferences toward non-meritocratic advantage. In other words, the classes differ less in whether they perceive privilege and more in whether they consider such privilege legitimate or acceptable.
```{r}
#| label: fig-prob
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "P(endorsement | class) for each item"
#| fig-cap-location: top

fit4 <- fits[[4]]

# 2) Extract conditional probabilities per item and class
# fit4$probs is a list: each element is K x Rj matrix.
# For dichotomous items: K x 2 (categories = colnames or "1","2")
probs_long <- bind_rows(lapply(names(fit4$probs), function(item) {
  mat <- fit4$probs[[item]]
  df <- as.data.frame(mat)
  df$class <- 1:nrow(mat)
  df$item <- item
  df
})) %>%
  pivot_longer(cols = -c(class, item), names_to = "category", values_to = "prob") %>%
  mutate(
    class = as.integer(class),
    prob = as.numeric(prob)
  )

# If levels are c("0","1") or c("1","2") you'll want the "higher"/"yes" one.
# Here I will assume endorsement is the LAST level (most common after dichotomization).
endorsement_level <- tail(levels(db_lca$a1), 1)

# 4) Build a clean table: P(endorsement | class) for each item
p_endorse <- probs_long %>%
  filter(category == "Pr(2)") %>%
  dplyr::select(item, class, p = prob) %>%
  mutate(item = factor(item, levels = names(fit4$probs)))

# 5) Plot: profile lines (one line per class)
p_endorse <- p_endorse %>% 
  mutate(
    variable = case_when(item == "a1" ~ "perc_effort",
                         item == "a2" ~ "perc_talent",
                         item == "a3" ~ "perc_rich_parents",
                         item == "a4" ~ "perc_contacts",
                         item == "a5" ~ "pref_effort",
                         item == "a6" ~ "pref_talent",
                         item == "a7" ~ "pref_rich_parents",
                         item == "a8" ~ "pref_contacts"
                         ),
    variable = factor(variable,
                      levels = c("perc_effort",
                                 "perc_talent",
                                 "perc_rich_parents",
                                 "perc_contacts",
                                 "pref_effort",
                                 "pref_talent",
                                 "pref_rich_parents",
                                 "pref_contacts")))

p_endorse <- p_endorse %>% 
  mutate(class = factor(class),
         p = round(p,2))

g_class <- ggplot(p_endorse, aes(x = variable, y = p, group = class, colour = class)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +  
  MetBrewer::scale_color_met_d("VanGogh2") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Item", y = paste0("P(Y = ", "High", " | class)"), group = "Class", colour = "Class") +
  theme(legend.position = "top",
        text = element_text(size = 14),
        axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1))


library(plotly)

ggplotly(g_class)
```


```{r}
#| label: tbl-prevalence
#| tbl-cap: "Items x classes with probabilities"
#| tbl-cap-location: top

# 7) Report table (wide): items x classes with probabilities
class_prev <- tibble(
  class = 1:length(fit4$P),
  pi = as.numeric(fit4$P)
) %>%
  mutate(pi_pct = 100 * pi)

report_probs <- p_endorse %>%
  mutate(class = paste0("Class_", class)) %>%
  pivot_wider(names_from = class, values_from = p) %>%
  mutate(across(starts_with("Class_"), ~ round(.x, 3)))

prev_lab <- class_prev %>%
  transmute(class = paste0("Class_", class),
            lab = paste0(class, " (", round(pi_pct, 1), "%)"))

report_probs2 <- report_probs
for (i in seq_len(nrow(prev_lab))) {
  old <- prev_lab$class[i]
  new <- prev_lab$lab[i]
  names(report_probs2)[names(report_probs2) == old] <- new
}

report_probs2[,-c(1)] %>% 
   kableExtra::kable(format = "html",
                    align = "c",
                    booktabs = T,
                    escape = F,
                    caption = NULL) %>%
  kableExtra::kable_styling(full_width = T,
                            latex_options = "hold_position",
                            bootstrap_options=c("striped", "bordered", "condensed"),
                            font_size = 23)
```



**Class 1: Legitimation of Privilege under Structural Realism (36.3%)**

The first class is characterized by low perceived meritocracy and very high perceived non-meritocracy. Respondents in this group do not believe that effort or talent are strongly rewarded, but they do strongly believe that family wealth and social contacts shape life chances. At the same time, they express high support for rewarding effort and talent, yet also show high normative acceptance of advantages linked to wealthy parents and personal networks.

Substantively, this class combines a realistic diagnosis of a socially stratified system with a normative legitimation of privilege. These respondents do not merely recognize that success is shaped by inherited or relational advantages; they also appear to accept that this is appropriate. Importantly, their endorsement of meritocratic rewards does not operate as a moral alternative to privilege, but rather coexists with it. This suggests a dual normative orientation in which merit and privilege are not seen as contradictory principles, but as compatible bases of social inequality.

**Class 2: Structurally Aware Meritocrats (20.7%)**

The second class displays very high levels of both perceived meritocracy and perceived non-meritocracy. Respondents in this group strongly believe that effort and talent are rewarded, while also recognizing the importance of wealthy parents and social contacts. Normatively, they express very strong support for meritocratic rewards, but only moderate acceptance of non-meritocratic advantages.

This profile suggests a belief system in which merit and privilege are understood as operating simultaneously. These respondents do not deny the role of structural advantage, yet neither do they reject the view that effort and talent matter. Normatively, they strongly endorse meritocracy, while showing a more ambivalent stance toward privilege. Compared to Class 1, they are less willing to legitimize inherited or network-based advantage, but they do not reject it entirely. This class can therefore be interpreted as one that sees the system as at least partially meritocratic and broadly considers this arrangement acceptable.

**Class 3: Weakly Structured or Ambivalent Orientations (6.8%)**

The third class is the smallest and least clearly structured. It is characterized by moderate levels of perceived meritocracy, low levels of perceived non-meritocracy, moderate support for meritocratic rewards, and relatively low support for non-meritocratic advantages. Unlike the other classes, respondents in this group do not strongly perceive either merit or privilege as dominant mechanisms of success.

Substantively, this class appears to represent a more ambivalent or weakly articulated orientation. Its members neither strongly endorse meritocratic ideals nor strongly perceive the system as dominated by privilege. Likewise, they do not express a strong normative defense of non-meritocratic advantage. Given its small size, this class should be interpreted with caution and evaluated for stability across specifications. Still, it may capture respondents whose views are less ideologically crystallized or whose understanding of success is not organized around the dominant opposition between merit and privilege.

**Class 4: Critical Meritocrats (36.2%)**

The fourth class combines low perceived meritocracy with very high perceived non-meritocracy. Respondents in this group do not believe that effort and talent are what society actually rewards, but they strongly believe that wealthy parents and social contacts shape success. Normatively, however, they strongly support meritocratic rewards and sharply reject the legitimacy of non-meritocratic advantages, especially those associated with personal networks.

This class is particularly important substantively because it captures a configuration in which meritocracy is upheld as a moral ideal while privilege is recognized as an empirical reality. In this sense, respondents in this group clearly distinguish between how society works and how it should work. They do not perceive the current system as meritocratic, but they do express strong support for a society in which effort and talent, rather than inherited or relational advantages, determine success. This class illustrates the analytical value of distinguishing perceptions from preferences, since older one-dimensional approaches to meritocratic beliefs would likely obscure this tension between normative commitment to merit and critical awareness of privilege.

**Overall Interpretation**

Taken together, the latent class solution suggests that meritocratic beliefs are best understood as multidimensional and internally heterogeneous. The main contrast is not between those who believe in meritocracy and those who do not, but between different ways of combining empirical perceptions of inequality with normative judgments about its legitimacy. Put differently, the classes vary across two central axes: first, whether respondents perceive success as driven by merit or by privilege; and second, whether they consider these mechanisms normatively acceptable.

This framework helps clarify the broader structure of the results. Class 2 combines recognition of both merit and privilege with strong support for meritocratic rewards and moderate tolerance of privilege. Class 4 combines recognition of privilege with a clear normative rejection of it, while maintaining strong support for merit as an ideal. Class 1 shares with Class 4 a critical diagnosis of how society works, but differs sharply in its normative legitimization of privilege. Finally, Class 3 stands apart as a smaller and less clearly structured group with comparatively weak commitments on both the perceptual and normative dimensions. Overall, the results show that support for meritocracy does not necessarily imply denial of privilege, and that recognition of privilege does not necessarily entail its rejection. This is precisely the kind of complexity that a person-centered latent class approach is able to recover.



```{r}
#| eval: false
#| include: false

fit4 <- fits[[4]]

post <- as.data.frame(fit4$posterior)
names(post) <- paste0("p_class", 1:4)

class_df <- na.omit(db_lca) %>%
  dplyr::select(id) %>%
  bind_cols(post) %>%
  mutate(
    class_hat = max.col(dplyr::select(., starts_with("p_class"))),
    class_hat = factor(class_hat)
  )

db_full <- db %>%
  left_join(class_df, by = "id")

db_full$just_pension <- as.numeric(db_full$just_pension)
db_full$just_pension <- if_else(db_full$just_pension >= 3, 1, 0)

P <- fits[[4]]$posterior
max_p <- apply(P, 1, max)
#summary(max_p)
#quantile(max_p, c(.1, .25, .5, .75, .9))

# MAP (ref=2)
db_reg1 <- db_full %>% filter(!is.na(class_hat), !is.na(just_pension), !is.na(age), !is.na(sex), !is.na(educ), !is.na(income_4))

db_reg1$class_hat <- relevel(db_reg1$class_hat, ref = "4")

m1 <- lm(just_pension ~ class_hat, data = db_reg1)
m2 <- lm(just_pension ~ class_hat + age, data = db_reg1)
m3 <- lm(just_pension ~ class_hat + age + sex, data = db_reg1)
m4 <- lm(just_pension ~ class_hat + age + sex + educ, data = db_reg1)
m5 <- lm(just_pension ~ class_hat + age + sex + educ + income_4, data = db_reg1)
m6 <- lm(just_pension ~ class_hat*income_4 + age + sex + educ, data = db_reg1)

# Posterior (ref implícita = clase 2)
#db_reg2 <- db_full %>% filter(!is.na(p_class2), #!is.na(just_pension))
#m_post <- lm(just_pension ~ p_class1 + p_class3 + p_class4, #data = db_reg2)
library(texreg)
screenreg(list(m1, m2, m2, m3, m4, m5))
screenreg(list(m6))
```


```{r}
#| label: fig-reg
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Market justice preferences in pensions by latent class of meritocracy (MAP model)"
#| fig-cap-location: top
#| eval: false
#| include: false

coef_df <-
  broom::tidy(m1, conf.int = TRUE) %>%
    mutate(model = "MAP (dummies; ref=Class 4)") %>%
    filter(term != "(Intercept)") %>%
    mutate(
      contrast = case_when(
        term == "class_hat1" ~ "Class 1 vs 4",
        term == "class_hat2" ~ "Class 2 vs 4",
        term == "class_hat3" ~ "Class 3 vs 4",
        TRUE ~ term
      )) %>%
  mutate(
    contrast = factor(contrast, levels = c("Class 1 vs 4", "Class 2 vs 4", "Class 3 vs 4"))
  )

ggplot(coef_df, aes(x = estimate, y = contrast)) +
  geom_vline(xintercept = 0, linewidth = 0.7, color = "grey60", linetype = "dashed") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), fatten = 2, 
                  size = 1, color = "#ca1137") + 
  labs(x = "Coefficient (OLS) with 95% CI", y = NULL) +
  theme(text = element_text(size = 14))
```


```{r}
#| label: fig-reg2
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Market justice preferences in pensions by latent class of meritocracy and covariates (MAP model)"
#| fig-cap-location: top
#| eval: false
#| include: false

coef_df <-
  broom::tidy(m5, conf.int = TRUE) %>%
    mutate(model = "MAP (dummies; ref=Class 4)") %>%
    filter(term != "(Intercept)") %>%
    mutate(
      contrast = case_when(
        term == "class_hat1" ~ "Class 1 vs 4",
        term == "class_hat2" ~ "Class 2 vs 4",
        term == "class_hat3" ~ "Class 3 vs 4",
        
        term == "sex2" ~ "Female (Ref.= Male)",
        term == "educ" ~ "Education (in years)",
        term == "age" ~ "Age",
        term == "income_4Medio-bajo" ~ "Income Middle-Low",
        term == "income_4Medio-alto" ~ "Income Middle-High",
        term == "income_4Alto" ~ "Income High",           
        TRUE ~ term
      )) 

ggplot(coef_df, aes(x = estimate, y = contrast)) +
  geom_vline(xintercept = 0, linewidth = 0.7, color = "grey60", linetype = "dashed") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), fatten = 2, 
                  size = 1, color = "#ca1137") + 
  labs(x = "Coefficient (OLS) with 95% CI", y = NULL) +
  theme(text = element_text(size = 14))
```

```{r}
#| label: fig-reg3
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Interaction between latent class of meritocracy and income on market justice preferences in pensions"
#| fig-cap-location: top
#| eval: false
#| include: false

coef_df <-
  broom::tidy(m6, conf.int = TRUE) %>%
    slice_tail(n=9) %>%
    mutate(
      contrast = case_when(
        term == "class_hat1" ~ "Class 1 vs 4",
        term == "class_hat2" ~ "Class 2 vs 4",
        term == "class_hat3" ~ "Class 3 vs 4",
        term == "sex2" ~ "Female (Ref.= Male)",
        term == "educ" ~ "Education (in years)",
        term == "age" ~ "Age",
        term == "income_4Medio-bajo" ~ "Income Middle-Low",
        term == "income_4Medio-alto" ~ "Income Middle-High",
        term == "income_4Alto" ~ "Income High",           
        TRUE ~ term
      )) 

ggplot(coef_df, aes(x = estimate, y = contrast)) +
  geom_vline(xintercept = 0, linewidth = 0.7, color = "grey60", linetype = "dashed") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), fatten = 2, 
                  size = 1, color = "#ca1137") + 
  labs(x = "Coefficient (OLS) with 95% CI", y = NULL) +
  theme(text = element_text(size = 14))
```


