---
title: "Data analysis"
subtitle: "Latent class analysis of the meritocracy scale"
author: "Andreas Laffert"
date: today
lang: en
fontsize: 12pt
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-location: right
    toc-depth: 2
    toc-expand: 2
    toc-title: Contents
    number-sections: true
    number-depth: 3
    theme:
      - cosmo
      - edumer_html.scss
    code-link: true
    title-block-banner: true
editor_options: 
  chunk_output_type: console
---

# Presentation

This is the analysis code for the future paper related to a latent class analysis of the meritocracy scale. The dataset used is `db_proc.RData` generated in the `prod_prep.R`  script.

```{r}
#| label: set
#| echo: false
#| message: false
#| warning: false

library(knitr)
knitr::opts_chunk$set(echo = TRUE, include = TRUE, warning = FALSE, message = FALSE)

table_format <- if(is_html_output()) {
  "html"
} else if(is_latex_output()) {
  "latex"
}
table_format2 <- if(is_html_output()) {
  T
} else if(is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```


# Libraries

```{r}
#| label: packages
#| include: true

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse,
               sjmisc, 
               here,
               lavaan,
               psych,
               corrplot,
               ggdist,
               sjlabelled,
               patchwork,
               RColorBrewer,
               poLCA,
               reshape,
               summarytools)

options(scipen=999)
rm(list = ls())
```

# Data

```{r}
#| label: data
#| echo: true

load(here("input/data/proc/db_proc.RData"))

glimpse(db)


```

# Analysis

## Descriptives



```{r}
#| label: tbl-summary1
#| tbl-cap: "Descriptive statistics"
#| tbl-cap-location: top



t1 <- db %>% 
   dplyr::select(c(3:6, 11:14, 7:10, 15:18))


df<-dfSummary(t1,
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               graph.magnif = 0.75,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = T,    # missing
               graph.col = T, # plot
               valid.col = T, # n valido
               col.widths = c(20,10,10,10,10,10))

df$Variable <- NULL # delete variable column

print(df, method="render")
```


```{r}
#| label: fig-likert
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Distribution of responses on the scale of perceptions and preferences regarding meritocracy"
#| fig-cap-location: top


vars_m <- c("perc_effort",
            "perc_talent",
            "perc_rich_parents",
            "perc_contact",
            "pref_effort",
            "pref_talent",
            "pref_rich_parents",
            "pref_contact")


db <- db %>% 
  mutate_all(~as.numeric(.))
 
labels1 <- c("Strongly desagree" = 1, 
             "Desagree" = 2, 
             "Agree" = 3, 
             "Strongly agree" = 4)
db <- db %>% 
  mutate(
    across(
      .cols = all_of(vars_m),
      .fns = ~  sjlabelled::set_labels(., labels = labels1)
    )
  )


df <- db %>% 
  dplyr::select(all_of(vars_m)) %>% 
  drop_na()

theme_set(theme_ggdist())
colors <- RColorBrewer::brewer.pal(n = 4, name = "RdBu")


a <- df %>% 
  dplyr::select(perc_effort, 
                perc_talent,
                perc_rich_parents,
                perc_contact) %>% 
  sjPlot::plot_likert(geom.colors = colors,
                      title = c("a. Perceptions"),
                      geom.size = 0.8,
                      axis.labels = c("Effort", "Talent", "Rich parents", "Contacts"),
                      catcount = 4,
                      values  =  "sum.outside",
                      reverse.colors = F,
                      reverse.scale = T,
                      show.n = FALSE,
                      show.prc.sign = T
  ) +
  ggplot2::theme(legend.position = "none")

b <- df %>% 
  dplyr::select(pref_effort, 
                pref_talent,
                pref_rich_parents,
                pref_contact) %>% 
  sjPlot::plot_likert(geom.colors = colors,
                      title = c("b. Preferences"),
                      geom.size = 0.8,
                      axis.labels = c("Effort", "Talent", "Rich parents", "Contacts"),
                      catcount = 4,
                      values  =  "sum.outside",
                      reverse.colors = F,
                      reverse.scale = T,
                      show.n = FALSE,
                      show.prc.sign = T
  ) +
  ggplot2::theme(legend.position = "bottom")

likerplot <- a / b + plot_annotation(caption = paste0("Source: own elaboration based on Survey EDUMERCO"," (n = ",dim(df)[1],")"
))

likerplot

```

```{r}
#| label: fig-cor
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Correlation matrix"
#| fig-cap-location: top

M <- df %>% 
  psych::polychoric()

diag(M$rho) <- NA

rownames(M$rho) <- c("A. Perception Effort",
                     "B. Perception Talent",
                     "C. Perception Rich parents",
                     "D. Perception Contacts",
                     "E. Preference Effort",
                     "F. Preference Talent",
                     "G. Preference Rich parents",
                     "H. Preference Contacts")

#set Column names of the matrix
colnames(M$rho) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)",
                    "(H)")

testp <- cor.mtest(M$rho, conf.level = 0.95)

#Plot the matrix using corrplot
corrplot::corrplot(M$rho,
                   method = "color",
                   addCoef.col = "black",
                   type = "upper",
                   tl.col = "black",
                   col = colorRampPalette(c("#E16462", "white", "#0D0887"))(12),
                   bg = "white",
                   na.label = "-") 

```


## Latent class analysis with dichotomized items

```{r}
#| include: false
# --- helpers -------------------------------------------------------------

get_K <- function(fit) {
  # robust: number of classes = columns of posterior
  if (!is.null(fit$posterior)) return(as.integer(ncol(fit$posterior)))
  # fallback: rows of probs for first item
  if (!is.null(fit$probs) && length(fit$probs) > 0) return(as.integer(nrow(fit$probs[[1]])))
  stop("Cannot determine K from poLCA object (no posterior/probs).")
}

get_npar <- function(fit) {
  if (!is.null(fit$npar)) return(as.numeric(fit$npar))
  # fallback if npar missing (rare): compute from probs + class shares
  K <- get_K(fit)
  Rj <- vapply(fit$probs, function(m) if (is.matrix(m)) ncol(m) else NA_integer_, integer(1))
  if (anyNA(Rj)) stop("Cannot compute npar: missing category counts in fit$probs.")
  as.numeric(sum(K * (Rj - 1)) + (K - 1))
}

get_n_patterns_obs <- function(fit) {
  # poLCA usually stores predicted cell counts for observed patterns as a vector
  if (!is.null(fit$predcell)) return(as.integer(length(fit$predcell)))
  # fallback: sometimes observed pattern table exists
  if (!is.null(fit$observed) && is.matrix(fit$observed)) return(as.integer(nrow(fit$observed)))
  NA_integer_
}

n_patterns_from_data <- function(data_indicators) {
  # data_indicators = db_lca with only indicators (your a1..a8), already NA-omitted
  as.integer(nrow(dplyr::distinct(data_indicators)))
}


entropy_stats <- function(fit) {
  P <- fit$posterior
  P <- pmax(P, 1e-12)
  N <- nrow(P); K <- ncol(P)
  
  # relative entropy (0..1): 1 = best separation
  rel_entropy <- 1 - (-sum(P * log(P)) / (N * log(K)))
  
  class_hat <- max.col(P)
  max_p <- apply(P, 1, max)
  
  avg_diag <- vapply(1:K, function(k) {
    idx <- which(class_hat == k)
    if (length(idx) == 0) NA_real_ else mean(P[idx, k])
  }, numeric(1))
  
  tibble(
    entropy = rel_entropy,
    mean_max_p = mean(max_p),
    p10_max_p = unname(quantile(max_p, 0.10)),
    avg_diag_min = suppressWarnings(min(avg_diag, na.rm = TRUE)),
    avg_diag_mean = mean(avg_diag, na.rm = TRUE)
  )
}

min_class_share <- function(fit) {
  Ppost <- fit$posterior
  N <- nrow(Ppost); K <- ncol(Ppost)
  class_hat <- max.col(Ppost)
  
  tibble(
    min_class_prior = min(fit$P),
    min_class_map   = min(tabulate(class_hat, nbins = K)) / N
  )
}

```


```{r}
#| include: false
# 3.3 LCA: dichotomized items ----

a1 <- db$perc_effort_d
a2 <- db$perc_talent_d
a3 <- db$perc_rich_parents_d
a4 <- db$perc_contact_d
a5 <- db$pref_effort_d
a6 <- db$pref_talent_d 
a7 <- db$pref_rich_parents_d
a8 <- db$pref_contact_d
id <- db$id

db_lca <- data.frame(id, a1, a2, a3, a4, a5, a6, a7, a8) %>% na.omit() %>% as_tibble()

db_lca <- db_lca %>% 
  mutate_at(.vars = 2:9, ~as.factor(.))

f <-cbind(a1, a2, a3, a4, a5, a6, a7, a8)~1

fit_lca <- function(K, data, formula, nrep = 50, maxiter = 3000, seed = 123) {
  set.seed(seed)
  poLCA(formula, data = data, nclass = K,
        nrep = nrep, maxiter = maxiter,
        verbose = FALSE, graphs = FALSE)
}

Ks <- 1:5
#fits <- lapply(Ks, fit_lca, data = db_lca, formula = f, nrep = 30)
#save(fits, file = here("output/fits_models_lca.RData"))

load(file = here("output/fits_models_lca.RData"))

fit_tbl <- data.frame(
  K    = Ks,
  logLik = sapply(fits, `[[`, "llik"),
  AIC  = sapply(fits, `[[`, "aic"),
  BIC  = sapply(fits, `[[`, "bic"),
  Gsq  = sapply(fits, `[[`, "Gsq"),
  Chisq = sapply(fits, `[[`, "Chisq")
)

fit_tbl
```

```{r}
#| label: tbl-fit
#| tbl-cap: "Fit comparision of latent class models"
#| tbl-cap-location: top

# ---- build table (no list-cols) -----------------------------------------

N_used <- nrow(db_lca)

m_patterns <- n_patterns_from_data(db_lca)  # scalar


# Robust extractors (avoid list-columns)
get_K <- function(fit) as.integer(ncol(fit$posterior))  # classes = cols of posterior
get_ll  <- function(fit) as.numeric(fit$llik)
get_aic <- function(fit) as.numeric(fit$aic)
get_bic <- function(fit) as.numeric(fit$bic)
get_npar <- function(fit) {
  if (!is.null(fit$npar)) return(as.numeric(fit$npar))
  # fallback: compute from probs
  K <- get_K(fit)
  Rj <- vapply(fit$probs, function(m) ncol(m), integer(1))
  as.numeric(sum(K * (Rj - 1)) + (K - 1))
}

# Table of fit indices
fit_tbl <- tibble(
  K     = map_int(fits, get_K),
  N     = nrow(db_lca),
  logLik= map_dbl(fits, get_ll),
  npar  = map_dbl(fits, get_npar),
  AIC   = map_dbl(fits, get_aic),
  BIC   = map_dbl(fits, get_bic)
) %>%
  arrange(K) %>%
  mutate(
    # Differences relative to preceding (K-1) model
    dAIC = AIC - lag(AIC),
    dBIC = BIC - lag(BIC),
    
    # Likelihood-ratio test (K vs K-1)
    # Test statistic: 2*(LL_K - LL_{K-1}) ~ Chi-square with df = npar_K - npar_{K-1}
    # Note: in LCA, models are not strictly nested in the regular sense;
    # this LRT is widely used as a heuristic, but interpret with caution.
    LRT = 2 * (logLik - lag(logLik)),
    df_LRT = npar - lag(npar),
    p_LRT = ifelse(!is.na(LRT) & df_LRT > 0, pchisq(LRT, df = df_LRT, lower.tail = FALSE), NA_real_)
  )


fit_tbl_report <- fit_tbl %>%
  transmute(
    K, N,
    logLik = round(logLik, 1),
    npar = npar,
    AIC = round(AIC, 1),
    dAIC = round(dAIC, 1),
    BIC = round(BIC, 1),
    dBIC = round(dBIC, 1),
    LRT = round(LRT, 1),
    df_LRT = df_LRT,
    p_LRT = signif(p_LRT, 3)
  )

fit_tbl_report %>% 
 kableExtra::kable(format = "html",
                    align = "c",
                    booktabs = T,
                    escape = F,
                    caption = NULL) %>%
  kableExtra::kable_styling(full_width = T,
                            latex_options = "hold_position",
                            bootstrap_options=c("striped", "bordered", "condensed"),
                            font_size = 23) %>%
  kableExtra::column_spec(c(1,8), width = "3.5cm") %>% 
  kableExtra::column_spec(2:7, width = "4cm") %>% 
  kableExtra::column_spec(4, width = "5cm")

```

```{r}
#| label: fig-visaic
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Visual: AIC/BIC by K (lower is better)"
#| fig-cap-location: top


# Visual: AIC/BIC by K (lower is better)
fit_long <- fit_tbl %>%
  dplyr::select(K, AIC, BIC) %>%
  pivot_longer(-K, names_to = "criterion", values_to = "value")

ggplot(fit_long, aes(x = K, y = value, group = criterion)) +
  geom_line() +
  geom_point() +
  facet_wrap(~criterion, scales = "free_y") +
  labs(x = "Number of classes (K)", y = "Value") 


```

```{r}
#| label: fig-dfvisaic
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "ΔAIC and ΔBIC (negative = improvement over K-1)"
#| fig-cap-location: top

# Visual: ΔAIC and ΔBIC (negative = improvement over K-1)
delta_long <- fit_tbl %>%
  dplyr::select(K, dAIC, dBIC) %>%
  pivot_longer(-K, names_to = "delta", values_to = "value")

ggplot(delta_long, aes(x = K, y = value, group = delta)) +
  geom_hline(yintercept = 0) +
  geom_line() +
  geom_point() +
  facet_wrap(~delta, scales = "free_y") +
  labs(x = "K", y = "Change vs K-1 (negative = better)") 
```



## Four latent class model


```{r}
#| label: fig-prob
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "P(endorsement | class) for each item"
#| fig-cap-location: top

fit4 <- fits[[4]]

# 2) Extract conditional probabilities per item and class
# fit4$probs is a list: each element is K x Rj matrix.
# For dichotomous items: K x 2 (categories = colnames or "1","2")
probs_long <- bind_rows(lapply(names(fit4$probs), function(item) {
  mat <- fit4$probs[[item]]
  df <- as.data.frame(mat)
  df$class <- 1:nrow(mat)
  df$item <- item
  df
})) %>%
  pivot_longer(cols = -c(class, item), names_to = "category", values_to = "prob") %>%
  mutate(
    class = as.integer(class),
    prob = as.numeric(prob)
  )


# If levels are c("0","1") or c("1","2") you'll want the "higher"/"yes" one.
# Here I will assume endorsement is the LAST level (most common after dichotomization).
endorsement_level <- tail(levels(db_lca$a1), 1)

# 4) Build a clean table: P(endorsement | class) for each item
p_endorse <- probs_long %>%
  filter(category == endorsement_level) %>%
  dplyr::select(item, class, p = prob) %>%
  mutate(item = factor(item, levels = names(fit4$probs)))

# 5) Plot: profile lines (one line per class)
p_endorse <- p_endorse %>% 
  mutate(
    variable = case_when(item == "a1" ~ "perc_effort",
                         item == "a2" ~ "perc_talent",
                         item == "a3" ~ "perc_rich_parents",
                         item == "a4" ~ "perc_contacts",
                         item == "a5" ~ "pref_effort",
                         item == "a6" ~ "pref_talent",
                         item == "a7" ~ "pref_rich_parents",
                         item == "a8" ~ "pref_contacts"
                         ),
    variable = factor(variable,
                      levels = c("perc_effort",
"perc_talent",
"perc_rich_parents",
"perc_contacts",
"pref_effort",
"pref_talent",
"pref_rich_parents",
"pref_contacts")))

ggplot(p_endorse, aes(x = variable, y = p, group = factor(class), colour = factor(class))) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Item", y = paste0("P(Y = ", "High", " | class)"), group = "Class", colour = "Class") +
  theme(legend.position = "bottom",
        text = element_text(size = 14),
        axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1))

```

```{r}
#| label: fig-heat
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Heatmap of endorsement"
#| fig-cap-location: top

# 6) Plot: heatmap (good for papers)
ggplot(p_endorse, aes(x = variable, y = factor(class), fill = p)) +
  geom_tile() +
  coord_cartesian(xlim = NULL) +
  labs(x = "Item", y = "Class", fill = paste0("P(Y=", "High", ")")) +
  theme(legend.position = "right",
        text = element_text(size = 14),
        axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1))
```

```{r}
#| label: tbl-prevalence
#| tbl-cap: "Items x classes with probabilities"
#| tbl-cap-location: top

# 7) Report table (wide): items x classes with probabilities
class_prev <- tibble(
  class = 1:length(fit4$P),
  pi = as.numeric(fit4$P)
) %>%
  mutate(pi_pct = 100 * pi)

report_probs <- p_endorse %>%
  mutate(class = paste0("Class_", class)) %>%
  pivot_wider(names_from = class, values_from = p) %>%
  mutate(across(starts_with("Class_"), ~ round(.x, 3)))

prev_lab <- class_prev %>%
  transmute(class = paste0("Class_", class),
            lab = paste0(class, " (", round(pi_pct, 1), "%)"))

report_probs2 <- report_probs
for (i in seq_len(nrow(prev_lab))) {
  old <- prev_lab$class[i]
  new <- prev_lab$lab[i]
  names(report_probs2)[names(report_probs2) == old] <- new
}

report_probs2[,-c(1)] %>% 
   kableExtra::kable(format = "html",
                    align = "c",
                    booktabs = T,
                    escape = F,
                    caption = NULL) %>%
  kableExtra::kable_styling(full_width = T,
                            latex_options = "hold_position",
                            bootstrap_options=c("striped", "bordered", "condensed"),
                            font_size = 23)
```

The @tbl-prevalence shows the conditional probabilities of each item associated with a latent class and their respective sizes. Class one has a size of 36.3% and shows low probabilities of perceiving meritocratic norms in society, while it highly perceives non-meritocratic functions. In addition, there is a high preference for both meritocratic and non-meritocratic elements, with talent being the only factor that deviates slightly from this pattern. In short, this class does not perceive a meritocratic functioning in society, and at the same time adheres to both meritocratic and non-meritocratic factors. This class is referred to as “**Anything goes preferences**.”

In the second class, there is a high perception of meritocratic elements and conditions of origin at the societal level. On the other hand, there is a strong preference for effort and talent, which differs from a lower adherence to non-meritocratic rules. In this case, there is a panoramic view of how society works, where effort and talent coexist with non-meritocratic functions; however, people prefer to be rewarded primarily for their merit. This group includes “**Meritocratic believers with a broad perspective**.”

The fourth class shows almost no perception of meritocratic norms in society, while perceiving a social order mainly governed by conditions of origin. In relation to this, preferences show a high adherence to the idea that society should function on the basis of meritocratic elements (although effort rather than talent), but having rich parents and good contacts should not influence the distribution of rewards in society. In this context, there is a clear narrative where, perceiving that society operates on the basis of non-meritocratic elements, there is a preference for meritocracy to come into play. This class is designated as “**Radical meritocrats**.”

# Regression anlaysis

```{r}
#| label: fig-reg
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Market justice preferences in pensions by latent class of meritocracy (MAP model)"
#| fig-cap-location: top

fit4 <- fits[[4]]

post <- as.data.frame(fit4$posterior)
names(post) <- paste0("p_class", 1:4)

class_df <- db_lca %>%
  dplyr::select(id) %>%
  bind_cols(post) %>%
  mutate(
    class_hat = max.col(dplyr::select(., starts_with("p_class"))),
    class_hat = factor(class_hat)
  )

db_full <- db %>%
  left_join(class_df, by = "id")

P <- fits[[4]]$posterior
max_p <- apply(P, 1, max)
#summary(max_p)
#quantile(max_p, c(.1, .25, .5, .75, .9))

# MAP (ref=2)
db_reg1 <- db_full %>% filter(!is.na(class_hat), !is.na(just_pension))
db_reg1$class_hat <- relevel(db_reg1$class_hat, ref = "2")
m_map <- lm(just_pension ~ class_hat, data = db_reg1)

# Posterior (ref implícita = clase 2)
db_reg2 <- db_full %>% filter(!is.na(p_class2), !is.na(just_pension))
m_post <- lm(just_pension ~ p_class1 + p_class3 + p_class4, data = db_reg2)

coef_df <-
  broom::tidy(m_map, conf.int = TRUE) %>%
    mutate(model = "MAP (dummies; ref=Class 4)") %>%
    filter(term != "(Intercept)") %>%
    mutate(
      contrast = case_when(
        term == "class_hat1" ~ "Class 1 vs 2",
        term == "class_hat3" ~ "Class 3 vs 2",
        term == "class_hat4" ~ "Class 4 vs 2",
        TRUE ~ term
      )) %>%
  mutate(
    contrast = factor(contrast, levels = c("Class 1 vs 2", "Class 3 vs 2", "Class 4 vs 2"))
  )

ggplot(coef_df, aes(x = estimate, y = contrast)) +
  geom_vline(xintercept = 0, linewidth = 0.7, color = "grey60", linetype = "dashed") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), fatten = 2, 
                  size = 1, color = "#ca1137") + 
  labs(x = "Coefficient (OLS) with 95% CI", y = NULL) +
  theme(text = element_text(size = 14))
```

